---
title: "ppgt: Pooled Group Testing Design Toolkit"
format: gfm
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[![License: GPL-3](https://img.shields.io/badge/License-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
[![CRAN status](https://www.r-pkg.org/badges/version/ppgt)](https://CRAN.R-project.org/package=ppgt)
[![](https://cranlogs.r-pkg.org/badges/grand-total/ppgt)](https://cran.r-project.org/package=ppgt)

## Overview

**ppgt** is an R package for constructing, analyzing, and comparing pooled
group testing designs. It brings together a set of classical and modern
designs for research and experimentation, including polynomial pooling,
PBEST, HYPER, Kirkman triple systems, other combinatorial designs (BIBDs,
hypercubes, arrays, projective planes), and separable/disjunct constructions,
together with finite-field (Galois field) utilities and COMP/GPSR decoders.

## Installation

`ppgt` is not yet on CRAN. Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("myaseen208/ppgt")

# or with remotes
# install.packages("remotes")
remotes::install_github("myaseen208/ppgt")
```

## Quick start

```{r}
library(ppgt)

# Construct a Polynomial Pools (PP) matrix -- the standard PP-based
# P-BEST configuration (q = 8, d = 3, nl = 6, N = 384)
M <- pp_matrix(q = 8, d = 3, nl = 6, N = 384)
dim(M)

# Simulate two positive individuals
x_true <- rep(0L, 384)
x_true[c(72, 142)] <- 1L

# Generate pooled outcomes and decode
z <- as.integer(as.vector(M %*% x_true) > 0)
result <- pp_decode(M, z)
result$positives
```

## Kirkman triple system (KTS) helpers

`kirkman_matrix(v)` constructs a resolvable Steiner triple system on `v`
points -- a Kirkman triple system exists exactly when `v %% 6 == 3`. Every
returned matrix is independently re-verified with `is_kts()` before it is
returned, checking all four properties that define a KTS: block size 3,
correct block count, every pair of points covered exactly once, and
resolvability into `(v-1)/2` parallel classes that each partition the
points:

```{r}
# v = 15 is the classic Kirkman schoolgirl problem
M15 <- kirkman_matrix(15)
dim(M15)                                  # 35 triples x 15 points
is_kts(M15)                               # verified: TRUE
attr(M15, "design")$parallel_class        # which of the 7 "days" each triple is in
```

Any `v` that is a power of 3 (`3, 9, 27, 81, ...`, via the affine geometry
AG(n,3)) and `v = 15` (the classical solution built from the lines of
PG(3,2)) construct quickly and deterministically. Other admissible `v`
(e.g. `21`, `33`, `63`) fall back to a general randomized search that is not
tuned for larger `v` -- it is not guaranteed to succeed even for `v = 21` --
in which case `kirkman_matrix()` raises an informative error rather than
ever returning an unverified result.

## Package contents

### Core designs

- `pp_matrix()` / `pp_decode()` / `pp_design()` / `pp_verify()` -- Polynomial
  Pools construction, decoding, and design search
- `hyper_matrix()` / `HyperDesign()` -- HYPER-style pooling matrices
- `hyper_ec_matrix()` / `hyper_ec_decode()` -- package-specific experimental
  parity-augmented extension of HYPER
- `tapestry_matrix()` -- Tapestry-style pooling matrix
- `SeparableMatrix()` / `DisjunctMatrix()` -- d-separable / d-disjunct
  constructions, self-verified against their defining properties
- `dorfman_matrix()` / `dorfman_optimal_g()` -- classic two-stage Dorfman
  pooling
- `array_matrix()` / `array_matrix_3d()` -- row/column and 3D array designs
- `bibd_matrix()` -- balanced incomplete block designs (built-in for
  specific `(v, k, lambda)`)
- `hypercube_matrix()` -- d-dimensional hypercube designs
- `kirkman_matrix()` / `is_kts()` -- Kirkman triple systems, see above
- `pg_matrix()` -- projective plane PG(2, q) designs
- `pbest_clinical_matrix()` -- clinical 384-to-94 P-BEST heuristic layout
- `regular_pooling_matrix()` / `score_pooling_matrix()` -- prescribed-marginal
  regular pooling matrices

### Finite-field utilities

- `gf()`, `gf_add()`, `gf_mult()`, `gf_pow()` -- Galois field construction and
  arithmetic
- `is_prime()`, `is_prime_power()`

### Workflow and comparison

- `PoolMatrix()` -- unified constructor across all design families
- `simulate_group_testing()`, `run_testing_workflow()`, `protocol_summary()`
- `compare_designs()`, `list_designs()`, `compare_all_designs()`,
  `compare_all_designs_honest()`

## Documentation

- Package reference: [github.com/myaseen208/ppgt](https://github.com/myaseen208/ppgt)
- Notation registry: `ppgt_notations()`
- Vignettes: `ppgt_theory` (overview), and topic articles `ppgt_pp`,
  `ppgt_pp_188x282` (worked 188 x 282 pooling matrix example), `ppgt_pbest`,
  `ppgt_hyper`, `ppgt_hyperec`, `ppgt_galois_field`,
  `ppgt_separable_disjunct`, `ppgt_comparisons`, `ppgt_diagnostics`

## License

GPL-3

## Author and citation

Package author and maintainer: **Muhammad Yaseen**
(<myaseen208@gmail.com>).

**Authors/Contributors**: Christopher McMahan, Christopher Bilder, Joshua Tebbs, and
Pranta Das.

```
Yaseen, M., McMahan, C.,  Bilder, C., Tebbs, J. and Das, P. (2026). ppgt: Pooled Group Testing Design Toolkit.
R package version 0.1.2. https://github.com/myaseen208/ppgt
```
