---
title: "Incidence-Matrix Diagnostics"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Incidence-Matrix Diagnostics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

This vignette develops two matrix diagnostics that are useful across the whole
package:

- pool overlap through $\mathbf{M}\mathbf{M}^\top$
- sample co-occurrence through $\mathbf{M}^\top\mathbf{M}$

These diagnostics are presented as vignette material only. They are not new
exported package helpers.

They also sit outside the constructor, decoder, simulation, and workflow APIs.
The matrices
$\mathbf{M}\mathbf{M}^\top$ and $\mathbf{M}^\top\mathbf{M}$ are
interpretive summaries of a fixed design $\mathbf{M}$; they do not choose a
decoder, execute retesting, or implement stage-2 protocol logic.

# Architecture Boundary

Within the package layer model:

1. construction returns $\mathbf{M}$
2. decoding interprets $\mathbf{z}$ relative to $\mathbf{M}$
3. simulation generates $\widetilde{\mathbf{z}}$ and $\mathbf{z}$
4. workflow execution coordinates stage-1 and optional stage-2 logic
5. reporting summarizes stored workflow objects

The diagnostics in this article belong to none of those execution layers. They
are read-only summaries of a fixed incidence matrix.

# Pool-Overlap Matrix

For $\mathbf{M} \in \{0,1\}^{J \times N}$, the pool-overlap matrix is
$\mathbf{M}\mathbf{M}^\top = \operatorname{tcrossprod}(\mathbf{M})$.

The $(j, j')$ entry counts how many individuals belong to both
$\mathcal{P}_j$ and $\mathcal{P}_{j'}$. The diagonal entries are the
pool sizes $n_j$.

More explicitly,
$$
(\mathbf{M}\mathbf{M}^\top)_{jj'}
=
\sum_{i=1}^N M_{ji} M_{j'i}
=
|\mathcal{P}_j \cap \mathcal{P}_{j'}|.
$$

```{r pool-overlap}
M_pp <- pp_matrix(q = 4, d = 3, nl = 5)
M_pp

pool_overlap <- Matrix::tcrossprod(M_pp)
pool_overlap

diag(as.matrix(pool_overlap))
```

# Sample Co-Occurrence Matrix

The sample co-occurrence matrix is
$\mathbf{M}^\top \mathbf{M} = \operatorname{crossprod}(\mathbf{M})$.

The $(i, i')$ entry counts how many pools contain both individuals
$i$ and $i'$. The diagonal entries are the sample coverages
$w_i$.

Equivalently,
$$
(\mathbf{M}^\top \mathbf{M})_{ii'}
=
\sum_{j=1}^J M_{ji} M_{ji'}
=
|\mathcal{J}_i \cap \mathcal{J}_{i'}|.
$$

```{r sample-cooccurrence}
sample_overlap <- Matrix::crossprod(M_pp)
sample_overlap

diag(as.matrix(sample_overlap))[1:12]
```

# Gramian Identities And Design Interpretation

The two diagnostics can be read as exact Gramian identities:
$$
\mathbf{G}_{\mathrm{pool}}
=
\mathbf{M}\mathbf{M}^\top,
\qquad
\mathbf{G}_{\mathrm{sample}}
=
\mathbf{M}^\top\mathbf{M}.
$$
Their diagonal entries recover the canonical row and column weights,
$$
(\mathbf{G}_{\mathrm{pool}})_{jj} = n_j,
\qquad
(\mathbf{G}_{\mathrm{sample}})_{ii} = w_i,
$$
while their off-diagonal entries recover pairwise overlaps,
$$
(\mathbf{G}_{\mathrm{pool}})_{jj'} = |\mathcal{P}_j \cap \mathcal{P}_{j'}|,
\qquad
(\mathbf{G}_{\mathrm{sample}})_{ii'} = |\mathcal{J}_i \cap \mathcal{J}_{i'}|.
$$
This viewpoint is useful across all design families because it turns qualitative
balance and overlap questions into direct matrix identities.

# Why These Diagnostics Matter

These matrices help interpret several design properties.

- Balance:
  the diagonals of $\mathbf{M}\mathbf{M}^\top$ and
  $\mathbf{M}^\top\mathbf{M}$ show whether $n_j$ and $w_i$ are
  constant.
- Overlap control:
  the off-diagonal entries show how often pools or individuals intersect.
- Separable and disjunct behavior:
  small sample co-occurrence makes it easier for different active sets to
  induce distinguishable pooled outcomes, and makes it easier for an outside
  individual to retain a pool that is not covered by a bounded active set.

# Comparing A Full PP Design And A Truncated One

```{r diagnostics-truncation}
M_full <- pp_matrix(q = 4, d = 3, nl = 5)
M_trunc <- pp_matrix(q = 4, d = 3, nl = 5, N = 48)

M_full
M_trunc

diag(as.matrix(Matrix::tcrossprod(M_full)))
diag(as.matrix(Matrix::tcrossprod(M_trunc)))
```

The full design has constant pool size on the diagonal. The truncated design
changes the pool-overlap diagonal because the row sizes $n_j$ are no
longer all equal.

# A Separable/Disjunct Interpretation Example

```{r diagnostics-sep-disj}
M_disj <- DisjunctMatrix(N = 16L, d = 2L, method = "reed-solomon")
M_disj

sample_overlap_disj <- Matrix::crossprod(M_disj)
sample_overlap_disj
sample_overlap_disj[1:12, 1:12]
```

Reading the off-diagonal structure of $\mathbf{M}^\top\mathbf{M}$ helps
explain why a design can separate or isolate individuals under bounded active
set sizes.

# End-To-End Workflow Example

The diagnostics are not themselves part of the workflow layer, but they can be
used to inspect a design before simulating outcomes and running a decoder.

```{r diagnostics-workflow}
M_workflow <- PoolMatrix(family = "pp", q = 4, d = 3, nl = 5, N = 64)
M_workflow

Y_tilde_workflow <- rep(0L, 64L)
Y_tilde_workflow[c(2L, 13L)] <- 1L
Y_tilde_workflow

pool_overlap_workflow <- Matrix::tcrossprod(M_workflow)
pool_overlap_workflow

sample_overlap_workflow <- Matrix::crossprod(M_workflow)
sample_overlap_workflow[1:12, 1:12]

sim_workflow <- simulate_group_testing(M_workflow, Y_tilde_workflow)
sim_workflow
sim_workflow$z_tilde
sim_workflow$z

workflow_result <- run_testing_workflow(
  design = M_workflow,
  z = sim_workflow$z,
  workflow = "non_adaptive",
  decoder = "pp_decode",
  decoder_args = list(verbose = FALSE),
  stage2 = "none"
)
workflow_result
workflow_result$stage1
workflow_result$final_calls

workflow_summary <- protocol_summary(workflow_result)
workflow_summary
```

# References

- Tan, Y. H. I. (2020). Pooling matrix designs for group testing. *SIAM
  Undergraduate Research Online*, 13, 1-21.
- Kautz, W. H., & Singleton, R. C. (1964). Nonrandom binary superimposed
  codes. *IEEE Transactions on Information Theory*, 10(4), 363-377.
