---
title: "ppgt: Theory, Notation, and Package Overview"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{ppgt: Theory, Notation, and Package Overview}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

The `ppgt` package provides pooled group testing matrix constructors, decoders,
finite-field utilities, and comparison helpers. This introductory article
summarizes the package notation, organizes the design families covered in the
package, introduces the incidence-matrix diagnostics used throughout the
vignette suite, and closes with a short end-to-end example.

# Package Introduction

In the package, a pooling design is represented by a binary incidence matrix
$\mathbf{M} \in \{0,1\}^{J \times N}$ with $J$ pools and $N$
individuals. The major exported families are:

- Polynomial Pools and PP-based P-BEST matrices
- HYPER-style matrices
- the package-specific experimental HyperEC extension
- finite-field arithmetic utilities for $GF(q)$
- separable and disjunct constructions
- package comparison helpers

The vignettes are split by topic so that the package overview remains concise
and the longer mathematical exposition stays with the relevant design family.

Given latent individual statuses
$\widetilde{\mathbf{y}} = (\widetilde{y}_1, \ldots, \widetilde{y}_N)^\top$,
the package uses the pooled binary status map
$$
\widetilde{z}_j =
\mathbb{I}\left(\sum_{i = 1}^N M_{ji}\widetilde{y}_i > 0\right),
\qquad
j \in \{1, \ldots, J\},
$$
with latent pool-status vector
$\widetilde{\mathbf{z}} = (\widetilde{z}_1, \ldots, \widetilde{z}_J)^\top$
and observed pool outcomes
$\mathbf{z} = (z_1, \ldots, z_J)^\top$.

# Package Layers

The package architecture is intentionally layered.

1. Matrix construction:
   `pp_matrix()`, `hyper_matrix()`, `SeparableMatrix()`, `DisjunctMatrix()`,
   the additional design-family constructors, and `PoolMatrix()` return a
   pooling design matrix $\mathbf{M}$ or a design-construction object.
2. Decoding:
   functions such as `pp_comp()`, `pp_gpsr()`, `pp_decode()`, and
   `hyper_ec_decode()` interpret observed outcomes relative to a fixed design.
3. Simulation:
   `simulate_group_testing()` maps a design and
   $\widetilde{\mathbf{y}}$ to latent and observed pool outcomes.
4. Workflow execution:
   `run_testing_workflow()` coordinates the currently implemented workflow
   skeleton above the constructor layer. In the present release, that means
   either stage-1 only (`workflow = "non_adaptive"`, `stage2 = "none"`) or
   stage-1 followed by individual retesting of the stage-1 non-negative set
   (`workflow = "adaptive_retest"`,
   `stage2 = "individual_retest"`).
5. Reporting:
   `protocol_summary()` summarizes a workflow object without constructing a
   matrix or re-running a decoder.

These layers are separate by design. In particular, `PoolMatrix()` remains a
constructor-only dispatcher that returns $\mathbf{M}$ and does not choose
adaptive versus non-adaptive workflows, retesting scenarios, stage-2 protocol
logic, or decoder behavior.

# Notation Summary

This package follows the notation defined in `Notations.tex`.

- $N$: total number of individuals
- $i \in \{1, \ldots, N\}$: individual index
- $\widetilde{\mathbf{y}} = (\widetilde{y}_1, \ldots, \widetilde{y}_N)^\top$:
  vector of latent infection indicators
- $J$: total number of pools
- $j \in \{1, \ldots, J\}$: pool index
- $\mathcal{P}_j$: set of individuals in pool $j$
- $n_j = |\mathcal{P}_j|$: size of pool $j$
- $\mathbf{M} = (M_{ji}) \in \{0,1\}^{J \times N}$: pooling design matrix
- $\mathcal{J}_i = \{j : M_{ji} = 1\}$: set of pools containing individual $i$
- $w_i = |\mathcal{J}_i|$: number of pools containing individual $i$
- $\widetilde{z}_j$: latent status of pool $j$
- $z_j$: observed test outcome of pool $j$
- $\widetilde{\mathbf{z}} = (\widetilde{z}_1, \ldots, \widetilde{z}_J)^\top$:
  vector of latent pool statuses
- $\mathbf{z} = (z_1, \ldots, z_J)^\top$: vector of observed pool results

For noiseless binary testing, the latent pool status is
$\widetilde{z}_j = \mathbb{I}\left(\sum_{i = 1}^N M_{ji} \widetilde{y}_i > 0\right)$.
Collecting the pool-level statuses gives
$\widetilde{\mathbf{z}} = (\widetilde{z}_1, \ldots, \widetilde{z}_J)^\top$.

## Package Notation Registry

The function `ppgt_notations()` exposes the package notation registry as a
plain data frame. It records the package-wide canonical notation, family-local
notation, decoder-local notation, display aliases, and deprecated aliases in
one place.

```{r notation-registry}
notations <- ppgt_notations()
notations

notations[
  notations$category == "global_canonical",
  c("symbol", "meaning", "family", "status")
]
```

The registry is intended to be the programmatic reference point for notation
used in the package documentation. Family-specific symbols such as PP
parameters or decoder-local symbols remain in the registry, but they are
explicitly marked as local rather than package-wide canonical notation.

# Package Organization

The vignette suite is organized as follows.

1. `ppgt_theory`: package introduction, notation, design-family tour, and a
   short workflow example
2. `ppgt_pp`: Polynomial Pools construction, truncation, and decoding
3. `ppgt_pbest`: PP-based P-BEST, later clinical P-BEST discussion, and
   package-specific boundaries
4. `ppgt_hyper`: HYPER matrix generation
5. `ppgt_hyperec`: package-specific experimental HyperEC extension
6. `ppgt_galois_field`: finite-field arithmetic
7. `ppgt_separable_disjunct`: separable and disjunct matrix constructors
8. `ppgt_comparisons`: package comparison helpers
9. `ppgt_diagnostics`: incidence-matrix diagnostics for overlap, balance, and
   interpretation

# Brief Tour Of The Design Families

The package spans several distinct construction families.

- Polynomial Pools:
  finite-field constructions indexed by $q$, $d$, and $n_l$, with
  sparse binary matrices produced by `pp_matrix()`.
- PP-based P-BEST:
  the standard package configuration `pp_matrix(q = 8, d = 3, nl = 6, N = 384)`
  corresponds to the 48-pool PP-based family used in the bundled P-BEST
  references.
- HYPER:
  `hyper_matrix()` is an upstream-compatible HYPER matrix generator for
  $q = 1$, $2$, or $3$.
- HyperEC:
  `hyper_ec_matrix()` and `hyper_ec_decode()` are package-specific
  experimental extensions layered on top of a base HYPER design.
- Separable and disjunct designs:
  `SeparableMatrix()` and `DisjunctMatrix()` provide random and
  Reed-Solomon-style paths, together with verification metadata.
- Comparisons:
  `compare_all_designs()` and `compare_all_designs_honest()` are package
  summaries of constructors, not literature results.

At a high level, the package covers both algebraic and combinatorial matrix
families. Polynomial Pools and the PP-based P-BEST family are driven by
finite-field arithmetic over $\mathrm{GF}(q)$. HYPER and HyperEC emphasize
balanced low-weight incidence patterns. The separable and disjunct
constructors target exact distinguishability properties for bounded active
sets. The additional design families expose classical screening layouts such as
Dorfman, arrays, BIBDs, hypercubes, Kirkman systems, and projective planes.

# Paper-Aligned, Compatible, And Package-Specific Scope

The package documentation distinguishes:

- paper-aligned construction claims, such as the PP construction and the
  standard PP-based P-BEST matrix family
- upstream-compatible implementation claims, such as the `hyper_matrix()`
  generator
- package-specific or experimental behavior, such as HyperEC and the summary
  labels used by the comparison helpers

# Workflow Boundary

Adaptive workflows, non-adaptive workflows, retesting scenarios, and
protocol-level decision rules belong above matrix construction.

- Use `PoolMatrix()` or a family-specific constructor to obtain
  $\mathbf{M}$.
- Use a decoder to interpret $\mathbf{z}$ relative to $\mathbf{M}$.
- Use `simulate_group_testing()` only when the task is to simulate
  $\widetilde{\mathbf{z}}$ and $\mathbf{z}$ from a fixed design.
- Use `run_testing_workflow()` only when the task is to coordinate the
  currently implemented stage-1 and optional stage-2 protocol logic.
- Use `protocol_summary()` only when the task is to summarize an existing
  workflow object.

This keeps matrix construction honest and keeps workflow behavior explicit.

# Constructor-Layer Example With `PoolMatrix()`

The constructor wrapper `PoolMatrix()` belongs only to the matrix-construction
layer. A simple deterministic PP example is enough to show that boundary
cleanly.

```{r poolmatrix-workflow}
M_pool <- PoolMatrix(family = "pp", q = 4, d = 3, nl = 5, N = 64)
M_pool

Y_tilde_pool <- rep(0L, 64)
Y_tilde_pool[c(2, 13)] <- 1L
Y_tilde_pool

sim_pool <- simulate_group_testing(M_pool, Y_tilde_pool)
sim_pool
sim_pool$z_tilde
sim_pool$z

wf_pool <- run_testing_workflow(
  design = M_pool,
  z = sim_pool$z,
  workflow = "non_adaptive",
  decoder = "pp_decode",
  decoder_args = list(verbose = FALSE),
  stage2 = "none"
)
wf_pool
wf_pool$stage1
wf_pool$final_calls

summary_pool <- protocol_summary(wf_pool)
summary_pool
```

In this example:

- `PoolMatrix()` returns only the stage-1 design matrix $\mathbf{M}$
- `simulate_group_testing()` generates $\widetilde{\mathbf{z}}$ and
  $\mathbf{z}$ from the fixed design
- `run_testing_workflow()` applies the currently implemented non-adaptive
  workflow skeleton above that design
- `protocol_summary()` reports the stored workflow result

The constructor itself still does not choose a workflow, perform retesting, or
run a decoder.

# Incidence-Matrix Diagnostics

Two matrix products are especially useful for reading a pooling design.

The pool-overlap matrix is
$\mathbf{M}\mathbf{M}^\top = \operatorname{tcrossprod}(\mathbf{M})$.

Its $(j, j')$ entry counts the number of individuals shared by pools
$\mathcal{P}_j$ and $\mathcal{P}_{j'}$. The diagonal entries are the
pool sizes $n_j$.

The sample co-occurrence matrix is
$\mathbf{M}^\top \mathbf{M} = \operatorname{crossprod}(\mathbf{M})$.

Its $(i, i')$ entry counts the number of pools shared by individuals
$i$ and $i'$. The diagonal entries are the sample coverages
$w_i$.

These two diagnostics help interpret:

- balance, through the diagonal structure
- overlap patterns, through the off-diagonal structure
- how strongly different pools or individuals are coupled
- why low shared coverage supports $d$-separable and
  $d$-disjunct interpretations

In particular, the diagonal entries recover the row and column weights
$$
(\mathbf{M}\mathbf{M}^\top)_{jj} = n_j = |\mathcal{P}_j|
\qquad\text{and}\qquad
(\mathbf{M}^\top\mathbf{M})_{ii} = w_i = |\mathcal{J}_i|,
$$
while the off-diagonal entries measure overlap:
$$
(\mathbf{M}\mathbf{M}^\top)_{jj'} = |\mathcal{P}_j \cap \mathcal{P}_{j'}|,
\qquad
(\mathbf{M}^\top\mathbf{M})_{ii'} = |\mathcal{J}_i \cap \mathcal{J}_{i'}|.
$$
These identities are useful across the entire package, regardless of whether
the design arises from PP, HYPER, a combinatorial block design, or a verified
separable/disjunct construction.

The full diagnostics discussion is developed in `ppgt_diagnostics`.

# Short End-To-End Example

```{r end-to-end}
M <- pp_matrix(q = 4, d = 3, nl = 5)
M

Y_tilde <- rep(0L, 64)
Y_tilde[c(2, 13)] <- 1L
Y_tilde

z <- as.integer(as.vector(M %*% Y_tilde) > 0)
z

decode <- pp_decode(M, z, verbose = FALSE)
decode

pool_overlap <- Matrix::tcrossprod(M)
pool_overlap

sample_overlap <- Matrix::crossprod(M)
sample_overlap[1:12, 1:12]
```

This short example shows the full package pattern:

1. build a matrix $\mathbf{M}$
2. simulate latent statuses $\widetilde{\mathbf{y}}$
3. compute observed pool results $\mathbf{z}$
4. decode
5. inspect overlap structure through `tcrossprod(M)`

# References

- Tan, Y. H. I. (2020). Pooling matrix designs for group testing. *SIAM
  Undergraduate Research Online*, 13, 1-21.
- Shental, N., Levy, S., Wuvshet, V., et al. (2020). Efficient high-throughput
  SARS-CoV-2 testing to detect asymptomatic carriers. *Science Advances*,
  6(37), eabc5961.
- Zismanov, V., Yelin, I., Klochendler, A., et al. (2024). High capacity
  clinical SARS-CoV-2 molecular testing using combinatorial pooling.
  *Communications Medicine*, 4, Article 121.
- Hong, F. S., Shah, N., Briggs, A., et al. (2022). HYPER group testing for
  SARS-CoV-2: A flexible, easy-to-implement, and highly efficient method.
  *Nature Communications*, 13, Article 3626.
