---
title: "GF(q) Arithmetic and Finite-Field Objects"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{GF(q) Arithmetic and Finite-Field Objects}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

The Polynomial Pools design family uses arithmetic in `GF(q)`, where `q` must
be a prime power. This vignette covers the package finite-field utilities:
`is_prime()`, `is_prime_power()`, `gf()`, `gf_add()`, `gf_mult()`, and
`gf_pow()`.

Let $q = p^n$, where $p$ is prime and $n \geq 1$. The package
finite-field constructor builds a
`galois_field` object encoding addition and multiplication in $GF(q)$.

For prime fields, arithmetic is ordinary modular arithmetic. For extension
fields, the package represents field elements by integer labels and uses lookup
tables induced by an irreducible polynomial model. Those lookup tables are what
the PP constructor ultimately uses when it evaluates polynomial relations over
$\mathrm{GF}(q)$.

# Prime And Prime-Power Checks

```{r primality}
is_prime(7)
is_prime(9)

pp8 <- is_prime_power(8)
pp8

pp10 <- is_prime_power(10)
pp10
```

The PP construction requires $q = p^n$ for a prime $p$ and an integer
$n \geq 1$.

# Creating `GF(q)`

```{r create-gf}
gf7 <- gf(7)
gf7

gf8 <- gf(8)
gf8
```

For `GF(7)`, the package uses modular arithmetic. For `GF(8)`, the package
uses an extension-field construction compatible with the package's finite-field
implementation choices for PP.

# Supported Fields

The exported `gf()` constructor supports:

- all prime fields $GF(p)$
- binary extensions through `GF(256)` for the implemented irreducible
  polynomials
- selected nonbinary extensions:
  `GF(9)`, `GF(25)`, `GF(27)`, and `GF(49)`

# Arithmetic Examples

```{r arithmetic}
gf_add(3, 5, gf7)
gf_mult(3, 5, gf7)

gf_add(5, 3, gf8)
gf_mult(2, 4, gf8)
gf_mult(5, 3, gf8)
gf_pow(3, 7, gf8)
```

These operations are used inside `pp_matrix()` when the construction evaluates
polynomials over `GF(q)`.

# Lookup Tables

The field objects also expose lookup tables. Printing a small slice is useful
for understanding the representation.

```{r lookup-tables}
gf8$add[1:4, 1:4]
gf8$mult[1:4, 1:4]
```

# Prime Field And Extension Field Contrast

```{r prime-vs-extension}
gf7$add[1:4, 1:4]
gf7$mult[1:4, 1:4]

gf9 <- gf(9)
gf9
gf9$add[1:4, 1:4]
gf9$mult[1:4, 1:4]
```

# Relationship To Polynomial Pools

In the PP construction, the matrix $\mathbf{M}$ is built by evaluating
field-valued polynomial expressions and then converting those evaluations into
binary pool assignments. The field arithmetic in this vignette is therefore the
algebraic foundation for `pp_matrix()`.

# References

- Lidl, R., & Niederreiter, H. (1997). *Finite Fields* (2nd ed.). Cambridge
  University Press.
- Tan, Y. H. I. (2020). Pooling matrix designs for group testing. *SIAM
  Undergraduate Research Online*, 13, 1-21.
