---
title: "HYPER: Matrix Generation and Interpretation"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{HYPER: Matrix Generation and Interpretation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

This vignette covers `hyper_matrix()` and `HyperDesign()`. In the package, the
HYPER implementation is documented as an upstream-compatible matrix generator.
It should be read as a construction for the stage-1 design matrix
$\mathbf{M}$, not as a claim that the package implements the full two-stage
protocol from the HYPER paper.

# Construction Summary

The package constructor returns a sparse matrix
$\mathbf{M} \in \{0,1\}^{J \times N}$ in which each individual belongs to
exactly $q$ pools, so $w_i = |\mathcal{J}_i| = q$.

The current implementation supports $q = 1$, $2$, and $3$, with
the same admissibility conditions documented in the exported help page.

At the design level, the HYPER generator is a balanced low-weight incidence
construction. The key exported claim is about the matrix
$\mathbf{M}$ itself: the package generates the stage-1 design and preserves the
upstream-compatible combinatorial factorization pattern for the supported
values of $q$.

# A Small HYPER Example

```{r hyper-small}
M_hyper <- hyper_matrix(n = 12, m = 6, q = 2, reorder = FALSE)
M_hyper

dim(M_hyper)
M_hyper[1:6, 1:12]
Matrix::colSums(M_hyper)
Matrix::rowSums(M_hyper)
```

For this example, every individual belongs to exactly `q = 2` pools, so
`w_i = |\mathcal{J}_i| = 2`.

# Reordering

```{r hyper-reorder}
M_hyper_reordered <- hyper_matrix(n = 12, m = 6, q = 2, reorder = TRUE)
M_hyper_reordered
```

The `reorder` argument changes only the column order of the generated factors.
It does not change the row space, the individual coverages, or the underlying
set of incidence relations.

# Compatibility Wrapper

```{r hyper-wrapper}
M_wrapper <- HyperDesign(n = 12, m = 6, q = 2, reorder = FALSE)
M_wrapper
```

`HyperDesign()` is a package wrapper around `hyper_matrix()`.

Both `hyper_matrix()` and `HyperDesign()` are matrix generators. They do not,
by themselves, implement the full HYPER workflow of stage-1 putative-positive
identification followed by stage-2 individual retesting.

In other words, this vignette is about the construction of
$\mathbf{M} \in \{0,1\}^{J \times N}$, not about a complete diagnostic
protocol layered above that matrix.

# Other Supported Cases

The package constructor supports `q = 1`, `q = 2`, and `q = 3`.

```{r hyper-q1-q3}
M_hyper_q1 <- hyper_matrix(n = 5, m = 5, q = 1, reorder = FALSE)
M_hyper_q1

M_hyper_q3 <- hyper_matrix(n = 24, m = 12, q = 3, reorder = FALSE)
M_hyper_q3

dim(M_hyper_q3)
```

# Scope Of The Package Claim

The package's claim for HYPER is compatibility at the matrix-generation level.
The HYPER paper itself studies a full testing procedure with stage-1 putative
positives and stage-2 retesting. This vignette stays at the matrix-construction
level, which is what `hyper_matrix()` implements.

# References

- Hong, F. S., Shah, N., Briggs, A., et al. (2022). HYPER group testing for
  SARS-CoV-2: A flexible, easy-to-implement, and highly efficient method.
  *Nature Communications*, 13, Article 3626.
- Upstream HYPER generator logic as summarized in the package documentation.
