---
title: "Package Comparison Helpers"
author: "Muhammad Yaseen"
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Package Comparison Helpers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

The functions `compare_all_designs()` and `compare_all_designs_honest()` are
package comparison helpers. They summarize matrices returned by package
constructors and attach package-specific labels. Their outputs should not be
read as direct literature results.

# Comparing Package Constructors At A Common Size

```{r compare-designs}
comparison <- compare_all_designs(n = 64, prevalence = 0.01, alpha = 0.20)
comparison

comparison$comparison_table
comparison$hyper$metrics
comparison$pbest$metrics
```

The comparison table reports structural summaries such as $J$, $N$,
$\lambda_{\max}$, and the ratio $J/N$.

# Reading The Metrics

For a design matrix $\mathbf{M} \in \{0,1\}^{J \times N}$, the helpers
summarize:

- $J$: number of pools
- $N$: number of individuals
- $J/N$: tests per individual in the design summary
- $\lambda_{\max} = \max_i |\mathcal{J}_i|$: maximum column weight

These summaries are intentionally structural. They compare the realized matrix
dimensions, row/column weights, and package reporting labels for the generated
design objects. They do not prove optimality, establish a universal ranking, or
replace family-specific theoretical guarantees.

# Comparing Package Reference Configurations

```{r compare-honest}
comparison_honest <- compare_all_designs_honest()
comparison_honest

comparison_honest$standard
comparison_honest$n256
comparison_honest$n384
comparison_honest$summary
```

# Interpreting The Labels

The package prints labels such as `Configuration`, `Error_Correction`, and the
entries in `summary`. Those labels are package-specific reporting choices for
these helpers. They are useful for package navigation, but they are not formal
paper-backed guarantees.

# References

- Shental, N., Levy, S., Wuvshet, V., et al. (2020). Efficient high-throughput
  SARS-CoV-2 testing to detect asymptomatic carriers. *Science Advances*,
  6(37), eabc5961.
- Ghosh, S., Agarwal, R., Rehan, M. A., et al. (2021). Tapestry: A
  single-round smart pooling technique for COVID-19 testing.
  *Nature Communications*, 12, Article 2995.
- Hong, F. S., Shah, N., Briggs, A., et al. (2022). HYPER group testing for
  SARS-CoV-2: A flexible, easy-to-implement, and highly efficient method.
  *Nature Communications*, 13, Article 3626.
