---
title: "P-BEST: PP-Based Designs and Clinical Context"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{P-BEST: PP-Based Designs and Clinical Context}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

This vignette distinguishes the standard PP-based 48-pool P-BEST family from
the later clinical P-BEST discussion. It also separates paper-aligned matrix
content from the package-specific helpers that live around the P-BEST design
family.

# The Standard PP-Based P-BEST Family

The standard PP-based P-BEST matrix family in the package is generated by
`pp_matrix(q = 8, d = 3, nl = 6, N = 384)`, which gives $J = 48$ pools
for $N = 384$ individuals.

At the matrix level, this is a PP design with
$$
q = 8, \qquad d = 3, \qquad n_l = 6, \qquad N = 384,
$$
so the full PP identity
$$
J = n_l q = 48
$$
matches the standard 48-pool family used in the package documentation.

# Reference-Aligned Matrix Example

```{r pbest-matrix}
M_pbest <- pp_matrix(q = 8, d = 3, nl = 6, N = 384)
M_pbest

dim(M_pbest)
attr(M_pbest, "pp")
```

This constructor call gives the package's PP-based 48-pool design family. The
key structural features are:

- $N = 384$ individuals
- $J = 48$ pools
- each individual appears in six pools
- each pool contains 48 individuals in the standard package configuration

# Structural Summaries

```{r pbest-structure}
Matrix::rowSums(M_pbest)[1:8]
Matrix::colSums(M_pbest)[1:8]
```

The displayed column sums show $w_i = 6$. The row sums show the realized
pool sizes $n_j$.

# Matrix-Family Alignment Versus Protocol Claims

The matrix call above is the package's reference-aligned claim for the
PP-based 48-pool family. That claim is narrower than saying that the package
reproduces every protocol, decoder, or laboratory workflow associated with
later P-BEST variants.

- Matrix-family alignment:
  the structural PP-based 48-pool construction represented by
  `pp_matrix(q = 8, d = 3, nl = 6, N = 384)`
- Package decoding:
  `pp_comp()` and `pp_decode()` are package decoding helpers built around the
  matrix interface
- Clinical package helpers:
  package-specific objects such as `pbest_clinical_matrix()` are documented as
  separate package functionality

This distinction matters because a pooling matrix is only one layer of a full
testing protocol. The matrix-level claim concerns the design
$\mathbf{M} \in \{0,1\}^{48 \times 384}$ and its PP structure. Laboratory
workflow rules, Ct-based interpretation, retesting policies, and stage-2
follow-up belong to different layers.

# Reference-Compatible Decoding Example

```{r pbest-decode}
Y_tilde_pbest <- rep(0L, 384)
Y_tilde_pbest[c(72, 142)] <- 1L
Y_tilde_pbest[c(72, 142)]

z_pbest <- as.integer(as.vector(M_pbest %*% Y_tilde_pbest) > 0)
z_pbest

comp_pbest <- pp_comp(M_pbest, z_pbest)
comp_pbest

decode_pbest <- pp_decode(M_pbest, z_pbest, verbose = FALSE)
decode_pbest

decode_pbest$positives
decode_pbest$suspected
```

# Distinguishing The 48-Pool Family From Later Clinical P-BEST

The 2020 P-BEST references and the later 2024 clinical P-BEST work should not
be conflated.

- The 48-pool PP-based family is the Reed-Solomon / PP-based configuration
  represented in the package by `pp_matrix(q = 8, d = 3, nl = 6, N = 384)`.
- The later clinical discussion concerns higher-capacity operational designs,
  workflow choices, Ct-based interpretation, and clinical deployment issues.

The package therefore keeps two separate stories:

- a PP-based matrix family story for the 48-pool algebraic design
- a clinical package-helper story for later operational high-capacity layouts

# Package-Specific Boundary

The matrix constructor above is the paper-aligned PP-based configuration. The
package also provides higher-level helpers around P-BEST-style workflows.

- `pp_decode()` is a package decoding pipeline.
- `pbest_clinical_matrix()` is a package clinical-design constructor.

These helpers are package functionality. They are not blanket claims that the
package reproduces every stage of each P-BEST workflow exactly.

# Package-Specific Clinical Constructor

```{r pbest-clinical}
clinical_design <- pbest_clinical_matrix()
clinical_design

dim(clinical_design)
attr(clinical_design, "design")
clinical_design[1:8, 1:16]
```

This object is documented separately from the 48-pool PP-based family so that
matrix-family alignment and package-specific clinical heuristics remain clearly
separated.

# References

- Shental, N., Levy, S., Wuvshet, V., et al. (2020). Efficient high-throughput
  SARS-CoV-2 testing to detect asymptomatic carriers. *Science Advances*,
  6(37), eabc5961.
- Zismanov, V., Yelin, I., Klochendler, A., et al. (2024). High capacity
  clinical SARS-CoV-2 molecular testing using combinatorial pooling.
  *Communications Medicine*, 4, Article 121.
