---
title: "Separable and Disjunct Matrix Families"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Separable and Disjunct Matrix Families}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

This vignette covers the exported constructors `SeparableMatrix()` and
`DisjunctMatrix()`.

A binary matrix $\mathbf{M} \in \{0,1\}^{J \times N}$ is
$d$-separable if every two distinct active sets
$\mathcal{S}_1, \mathcal{S}_2 \subseteq \{1, \ldots, N\}$ with
$|\mathcal{S}_1| \leq d$ and $|\mathcal{S}_2| \leq d$ induce distinct
pooled outcomes. Equivalently, the pooled outcome vector
$\widetilde{\mathbf{z}}$ determines the active set uniquely whenever at
most $d$ individuals are positive.

A matrix is $d$-disjunct if for every active set
$\mathcal{S} \subseteq \{1, \ldots, N\}$ with $|\mathcal{S}| \leq d$
and every outside individual $i^\star \notin \mathcal{S}$, there is at
least one pool containing $i^\star$ that contains none of the individuals
in $\mathcal{S}$. This is stronger than $d$-separability.

Using the pool index sets
$\mathcal{J}_i = \{j : M_{ji} = 1\}$, the $d$-disjunct condition can be written
as
$$
\mathcal{J}_{i^\star}
\nsubseteq
\bigcup_{i \in \mathcal{S}} \mathcal{J}_i
\qquad
\text{for every } |\mathcal{S}| \le d \text{ and } i^\star \notin \mathcal{S}.
$$
That is the structural reason a $d$-disjunct matrix permits exact elimination
of nondefective individuals from pooled positives.

# Union Maps And Distinguishability

For any active set $\mathcal{S} \subseteq \{1, \ldots, N\}$, define the induced
pool-status pattern
$$
\mathbf{u}_{\mathcal{S}}
=
\bigvee_{i \in \mathcal{S}} \mathbf{M}_{\cdot i}
\in \{0,1\}^J.
$$
This notation is local to the separable/disjunct theory and is recorded that
way in `ppgt_notations()`.

The two core notions can then be written compactly:

- $d$-separable means that
  $\mathbf{u}_{\mathcal{S}_1} \ne \mathbf{u}_{\mathcal{S}_2}$ whenever
  $\mathcal{S}_1 \ne \mathcal{S}_2$ and
  $|\mathcal{S}_1|, |\mathcal{S}_2| \le d$.
- $d$-disjunct means that for every
  $\mathcal{S} \subseteq \{1,\ldots,N\}$ with $|\mathcal{S}| \le d$ and every
  outside individual $i^\star \notin \mathcal{S}$,
  $\mathbf{M}_{\cdot i^\star} \nleq \mathbf{u}_{\mathcal{S}}$ entrywise.

So $d$-separability asks whether bounded active sets have distinct pooled
signatures, while $d$-disjunctness asks for a stronger one-vs-set isolation
property.

# Separable Construction Paths

`SeparableMatrix()` supports:

- an automatic path
- a random construction
- a Reed-Solomon-style construction

# A Separable Design

```{r sep-design}
M_sep <- SeparableMatrix(N = 16L, d = 2L, method = "reed-solomon")
M_sep

dim(M_sep)
M_sep[1:8, 1:16]
attr(M_sep, "method_used")
attr(M_sep, "separable_verified")
```

# Disjunct Construction Paths

`DisjunctMatrix()` supports the same broad path structure, but targets the
stronger $d$-disjunct property.

# A Disjunct Design

```{r disj-design}
M_disj <- DisjunctMatrix(N = 16L, d = 2L, method = "reed-solomon")
M_disj

dim(M_disj)
M_disj[1:8, 1:16]
attr(M_disj, "method_used")
attr(M_disj, "disjunct_verified")
```

# Relationship Between The Two

Every `d`-disjunct design is also `d`-separable, but the converse need not
hold.

The inclusion is strict in general. A $d$-separable design only requires
distinct active sets of size at most $d$ to induce distinct pooled outcomes,
whereas a $d$-disjunct design requires every outside individual to retain a
pool not covered by any active set of size at most $d$.

Equivalently, if
$$
\mathbf{u}_{\mathcal{S}}
=
\bigvee_{i \in \mathcal{S}} \mathbf{M}_{\cdot i},
$$
then $d$-disjunctness requires an outside individual $i^\star$ to have at least
one row with
$$
M_{j i^\star} = 1
\qquad\text{and}\qquad
u_{\mathcal{S}, j} = 0.
$$
That extra uncovered row is what COMP exploits when a disjunct design is used
as a decoder-friendly construction.

```{r compare-sep-disj}
Matrix::colSums(M_sep)
Matrix::colSums(M_disj)
```

# Overlap Diagnostics For Separable And Disjunct Designs

The sample co-occurrence matrix
$$
\mathbf{M}^\top \mathbf{M}
$$
helps explain why these constructions work. Its diagonal is the column-weight
vector $w_i$, and its off-diagonal entries measure how many pools are shared by
pairs of individuals. Small off-diagonal overlap leaves more rows available to
separate one active pattern from another.

```{r sep-disj-gramians}
sep_overlap <- Matrix::crossprod(M_sep)
sep_overlap[1:12, 1:12]

disj_overlap <- Matrix::crossprod(M_disj)
disj_overlap[1:12, 1:12]
```

# Verification Behavior

Both constructors attach verification metadata when verification is enabled.

```{r verification-metadata}
attributes(attr(M_sep, "separable_verified"))
attributes(attr(M_disj, "disjunct_verified"))
```

# Package Scope

The package also contains internal verification helpers for separability and
disjunctness. This vignette focuses on the exported constructors and their
returned sparse matrices.

# References

- Kautz, W. H., & Singleton, R. C. (1964). Nonrandom binary superimposed codes.
  *IEEE Transactions on Information Theory*, 10(4), 363-377.
- D'yachkov, A. G., & Rykov, V. V. (1982). Bounds on the length of disjunctive
  codes. *Problemy Peredachi Informatsii*, 18(3), 7-13.
