---
title: "Polynomial Pools: Construction, Truncation, and Decoding"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Polynomial Pools: Construction, Truncation, and Decoding}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

This vignette develops the Polynomial Pools (PP) construction implemented by
`pp_matrix()`. It describes the full design with $N = q^d$, the package
handling of truncated designs with $N < q^d$, and the public decoding
helpers `pp_comp()`, `pp_gpsr()`, `pp_decode()`, and `pp_validate()`.

# PP Construction

Let $\mathbf{M} \in \{0,1\}^{J \times N}$ denote a PP matrix. In the full
untruncated construction:

- $N = q^d$
- $J = n_l q$
- $w_i = |\mathcal{J}_i| = n_l$ for every individual $i$
- $n_j = |\mathcal{P}_j| = q^{d-1}$ for every pool $j$

The construction indexes individuals by field-valued coordinates and assigns
them to pools by polynomial evaluation over $GF(q)$. In the full design,
the number of pools is $J = n_l q$, and the standard COMP detection bound
is $k_{\max} = \left\lfloor (n_l - 1) / (d - 1) \right\rfloor$.

More explicitly, a PP design with field order $q$ and dimension $d$ starts
from the full ambient index set of size
$$
N_{\max} = q^d.
$$
Individuals are represented by field-valued coordinate tuples, and each layer
corresponds to a family of polynomial evaluations over $\mathrm{GF}(q)$. In
the full design,
$$
J = n_l q,
\qquad
w_i = n_l,
\qquad
n_j = q^{d-1},
\qquad
k_{\max} = \left\lfloor \frac{n_l - 1}{d - 1} \right\rfloor.
$$
The PP intersection bound implies that any two distinct individuals share at
most $d - 1$ pools.

## Why The COMP Bound Follows

The PP guarantee used throughout the package is a direct consequence of the
intersection bound and the uniform column weight $w_i = n_l$ in the full
design.

Let $i^\star$ be an uninfected individual and let
$\mathcal{S} \subseteq \{1, \ldots, N\}$ be the active set of truly infected
individuals, with $|\mathcal{S}| = k$. Since any two distinct columns of a PP
matrix meet in at most $d - 1$ pools, the pools containing $i^\star$ that are
also contaminated by positives are bounded by
$$
\left|
\mathcal{J}_{i^\star}
\cap
\bigcup_{i \in \mathcal{S}} \mathcal{J}_i
\right|

\le
k(d - 1).
$$
If
$$
n_l > k(d - 1),
$$
then at least one pool containing $i^\star$ remains negative, so COMP removes
$i^\star$ from the candidate set. Solving this inequality for the largest
guaranteed $k$ yields
$$
k_{\max}
=
\left\lfloor \frac{n_l - 1}{d - 1} \right\rfloor.
$$
That is the package-wide reason the PP help pages state exact COMP recovery for
active sets up to $k_{\max}$.

# Building A PP Matrix

```{r build-pp}
M_pp <- pp_matrix(q = 4, d = 3, nl = 5)
M_pp

dim(M_pp)
attr(M_pp, "pp")
```

This small example matches the standard `q = 4`, `d = 3`, `k = 2` PP family
used throughout the package documentation.

# Verifying Structure

```{r verify-pp}
pp_summary <- pp_verify(M_pp)
pp_summary
```

The helper checks the realized row sizes $n_j = |\mathcal{P}_j|$ and column
weights $w_i = |\mathcal{J}_i|$.

# Truncation Behavior

The constructor also supports $N < q^d$. The column weights stay fixed at
$w_i = n_l$, while row sizes near the truncation boundary need not stay
equal.

Operationally, truncation keeps the first $N$ columns of the full PP design and
therefore preserves the layer structure at the individual level. The cost is
that the uniform row identity $n_j = q^{d-1}$ is no longer guaranteed once some
columns are removed.

```{r truncation}
M_trunc <- pp_matrix(q = 4, d = 3, nl = 5, N = 48)
M_trunc

dim(M_trunc)
table(Matrix::rowSums(M_trunc))
table(Matrix::colSums(M_trunc))
M_trunc[1:8, 1:16]
```

This is the key structural distinction between the full PP design and a
truncated one.

## Matrix Products And PP Structure

Two Gramian-style products are especially informative for PP matrices:
$$
\mathbf{M}\mathbf{M}^\top
\qquad\text{and}\qquad
\mathbf{M}^\top\mathbf{M}.
$$
For the full design, their diagonals recover the canonical row and column
weights,
$$
(\mathbf{M}\mathbf{M}^\top)_{jj} = n_j = q^{d-1},
\qquad
(\mathbf{M}^\top\mathbf{M})_{ii} = w_i = n_l,
$$
while the off-diagonal entries record pairwise overlaps. The PP intersection
bound is a statement about the off-diagonal entries of
$\mathbf{M}^\top\mathbf{M}$:
$$
(\mathbf{M}^\top\mathbf{M})_{ii'}
=
|\mathcal{J}_i \cap \mathcal{J}_{i'}|
\le d - 1
\qquad
\text{for } i \ne i'.
$$
This is the matrix-product view of the same combinatorial structure used by the
COMP recovery argument.

```{r pp-gramians}
pp_pool_overlap <- Matrix::tcrossprod(M_pp)
pp_pool_overlap

pp_sample_overlap <- Matrix::crossprod(M_pp)
pp_sample_overlap[1:12, 1:12]
```

# Latent And Observed Pool Status

Given latent individual statuses
$\widetilde{\mathbf{y}} \in \{0,1\}^N$, the latent pool status is
$\widetilde{z}_j = \mathbb{I}\left(\sum_{i = 1}^N M_{ji}\widetilde{y}_i > 0\right)$.

In the simple noiseless examples below, we set the observed outcomes equal to
the latent pool statuses.

Equivalently,
$$
\widetilde{\mathbf{z}}
=
\left(
\mathbb{I}\left((\mathbf{M}\widetilde{\mathbf{y}})_1 > 0\right),
\ldots,
\mathbb{I}\left((\mathbf{M}\widetilde{\mathbf{y}})_J > 0\right)
\right)^\top,
$$
and the noiseless examples use $\mathbf{z} = \widetilde{\mathbf{z}}$.

```{r simulate-pp}
Y_tilde <- rep(0L, 64)
Y_tilde[c(2, 13)] <- 1L
Y_tilde

z <- as.integer(as.vector(M_pp %*% Y_tilde) > 0)
z
```

# COMP Decoding

```{r comp-pp}
comp_result <- pp_comp(M_pp, z)
comp_result
```

For this small example, `pp_comp()` recovers the two positives exactly.
The COMP rule marks an individual as negative whenever that individual appears
in at least one observed negative pool. If $\mathcal{N}_0 = \{j : Z_j = 0\}$
is the set of negative pools, then the COMP candidate set is
$$
\mathcal{C}_{\mathrm{COMP}}
=
\left\{
i \in \{1, \ldots, N\} :
M_{ji} = 0 \text{ for all } j \in \mathcal{N}_0
\right\}.
$$

# GPSR Candidate Scoring

`pp_gpsr()` solves the nonnegative $\ell_1$-regularized least-squares
problem used by the package sparse-scoring layer.

In this decoder-local subsection, the symbols $\mathbf{A}$, $\mathbf{x}$, and
$\mathbf{y}$ follow the sparse-reconstruction literature and are local to the
decoder. They do not replace the package-wide canonical symbols
$\mathbf{M}$, $\widetilde{\mathbf{y}}$, and $\mathbf{z}$. In that
decoder-local notation, the scoring problem is
$$
\min_{\mathbf{x} \ge 0}
\frac{1}{2}\|\mathbf{y} - \mathbf{A}\mathbf{x}\|_2^2
+ \tau \|\mathbf{x}\|_1,
$$
where $\mathbf{A}$ is typically the positive-pool submatrix restricted to the
COMP candidate set. The solution is interpreted as a sparse score vector over
those candidates rather than as a final binary call by itself.

```{r gpsr-pp}
A_sub <- M_pp[which(z == 1), comp_result$candidates, drop = FALSE]
y_sub <- rep(1, nrow(A_sub))
A_sub
y_sub

gpsr_result <- pp_gpsr(A_sub, y_sub)
gpsr_result
```

# Complete Package Decoding Pipeline

`pp_decode()` is the package pipeline combining COMP, verification, and GPSR
refinement when needed. It is package functionality built around the PP matrix
interface; it should not be described as a blanket claim of exact equivalence
to every bundled upstream decoding path.

The implemented logic is:

1. threshold or binarize the observed pool outcomes $\mathbf{z}$
2. run COMP to eliminate obvious negatives
3. if the candidate set is already small enough, return a direct stage-1 call
4. otherwise score the candidate set with GPSR
5. search a bounded sparse subset and report package decoding outputs such as
   `positive`, `suspected`, and `negative`

```{r decode-pp}
decode_result <- pp_decode(M_pp, z, verbose = FALSE)
decode_result

decode_result$positives
decode_result$suspected
decode_result$negatives[1:12]
```

# Validation Against Known Positives

```{r validate-pp}
validation <- pp_validate(decode_result$positives, c(2, 13))
validation
```

# Exploring Candidate Configurations

```{r pp-design}
design_grid <- pp_design(N_min = 50, N_max = 150, M_max = 40, k_min = 2, pool_size_max = 32)
design_grid
head(design_grid, 6)
```

# References

- Tan, Y. H. I. (2020). Pooling matrix designs for group testing. *SIAM
  Undergraduate Research Online*, 13, 1-21.
- PP reference algorithms as discussed in the package documentation and source
  comparisons.
