---
title: "HyperEC: Package-Specific Experimental Extension"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{HyperEC: Package-Specific Experimental Extension}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

# Abstract

This vignette covers the package-specific experimental HyperEC extension. The
functions `hyper_ec_matrix()` and `hyper_ec_decode()` are not the HYPER paper
design itself. They are documented as package-specific experimental tools built
on top of a base HYPER matrix.

# Scope And Limits

The current HyperEC interface combines:

1. a base HYPER matrix
2. additional parity-derived rows
3. a bounded decoder that searches for a low-weight correction pattern

This should be interpreted as a package experiment. It is not an exact
implementation of a Hong et al. HyperEC procedure.

The returned object contains a base HYPER matrix together with additional
parity-derived rows. If the combined incidence matrix is denoted by
$\mathbf{M}_{\mathrm{EC}}$, then the package reports
$$
\mathbf{M}_{\mathrm{EC}}
=
\begin{bmatrix}
\mathbf{M}_{\mathrm{base}} \\
\mathbf{M}_{\mathrm{parity}}
\end{bmatrix},
$$
but the exact parity-generation rule and the decoder interpretation are
package-specific rather than paper-backed claims.

# Building An Experimental HyperEC Design

```{r hyper-ec-build}
hyper_ec_design <- hyper_ec_matrix(n = 64, k = 2, q = 2, alpha = 0.20)
hyper_ec_design

dim(hyper_ec_design$matrix)
dim(hyper_ec_design$matrix_base)
dim(hyper_ec_design$matrix_parity)
```

The combined design contains a base matrix and additional parity-derived rows.

# Inspecting The Experimental Metadata

```{r hyper-ec-metadata}
hyper_ec_design$error_correction
hyper_ec_design$efficiency
range(Matrix::colSums(hyper_ec_design$matrix))
```

The stored `error_correction` value is a package search limit used by the
current decoder implementation. It should not be interpreted as a formal
literature-backed correction guarantee.

# A Small Decode Example

```{r hyper-ec-decode}
Y_tilde_hyper_ec <- rep(0L, 64)
Y_tilde_hyper_ec[c(3, 20)] <- 1L
Y_tilde_hyper_ec

z_hyper_ec <- as.integer(
  as.vector(hyper_ec_design$matrix %*% Y_tilde_hyper_ec) > 0
)
z_hyper_ec

decode_hyper_ec <- hyper_ec_decode(z_hyper_ec, hyper_ec_design)
decode_hyper_ec
```

# Interpretation

The object returned by `hyper_ec_decode()` is best read as a stage-1 candidate
screening result under the package's current parity model, not as a general
error-correction guarantee.

# Experimental Boundary

This vignette is intentionally separate from the HYPER vignette so that the
package-specific extension is not presented as paper-aligned HYPER behavior.

# References

- Hong, F. S., Shah, N., Briggs, A., et al. (2022). HYPER group testing for
  SARS-CoV-2: A flexible, easy-to-implement, and highly efficient method.
  *Nature Communications*, 13, Article 3626.
- Package-specific experimental implementation in `R/tapestry_hyperec.R`.
