---
title: "Building a 188 x 282 pooling matrix"
author: "Muhammad Yaseen"
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    toc: true
    toc-depth: 3
    highlight-style: github
    code-fold: false
    code-tools: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Building a 188 x 282 pooling matrix}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
if (requireNamespace("ppgt", quietly = TRUE)) {
  library(ppgt)
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
# ppgt pooling matrices are Matrix::dgCMatrix objects; attaching Matrix lets
# the base rowSums()/colSums() generics dispatch on them below.
library(Matrix)
```

# Purpose

This vignette works through a single concrete Polynomial Pools (PP) design:
**282 specimens tested using 188 pools**, where every specimen is placed in
exactly 4 pools and every pool holds exactly 6 specimens. That is a design
with

- $N = 282$ specimens (columns of $\mathbf{M}$),
- $J = 188$ pools (rows of $\mathbf{M}$),
- pool size $n_j = 6$ for every pool $j$,
- column weight $w_i = 4$ for every specimen $i$.

# Construction

The matrix is built with a single call to `pp_matrix()`:

```{r build}
M4 <- pp_matrix(q = 47, d = 3, nl = 4, N = 282)
dim(M4)
```

`pp_matrix()` builds its design over the finite field $\mathrm{GF}(q)$, using
dimension `d` and `nl` layers (see `?pp_matrix` and the `ppgt_pp` vignette for
the full construction). The three arguments interact as follows:

- `q = 47` is the field order. 47 is prime, so $\mathrm{GF}(47)$ exists, and
  each layer of the design contributes exactly $q = 47$ pools.
- `nl = 4` is the number of layers, so the pool count is
  $J = n_l \, q = 4 \times 47 = 188$, and every specimen is assigned to
  exactly $n_l = 4$ pools (one per layer).
- `d = 3` is the PP dimension, which fixes the canonical pool size to
  $q^{d-1} = 47^2 = 2209$ in the *full* (untruncated) design with
  $N_{\max} = q^d = 47^3$ specimens.
- `N = 282` truncates that full design down to its first 282 columns. Because
  $282 = 6 \times 47 = 6q$ is a small multiple of $q$, the truncated design
  keeps only 6 of the $q^{d-2} = 47$ possible "column blocks" per pool, so
  every pool's row sum drops from the full-design value of 2209 down to 6.

The two products $188 = 4 \times 47$ and $282 = 6 \times 47$ both factor
through $q = 47$, and this is exactly what keeps every row sum equal to 6 and
every column weight equal to 4 (rather than only approximately so): $q$
divides $N$ exactly (`pp_matrix()` requires `N %% q == 0`), so the truncation
boundary falls cleanly between pool blocks and does not cut any pool's row in
half.

Every incidence $M_{ji} = 1$ is counted once from the pool side (as a 1 in
row $j$) and once from the specimen side (as a 1 in column $i$), so the total
number of incidences can be computed two ways:
$$
J \, n_j = 188 \times 6 = 1128, \qquad N \, w_i = 282 \times 4 = 1128.
$$
Both give the same total, 1128, which is confirmed numerically below.

# Verification

```{r verify}
dim(M4)

# apply() works directly on the sparse matrix, matching the original script
table(apply(M4, 1, sum))
table(apply(M4, 2, sum))

# rowSums()/colSums() dispatch to Matrix's sparse-aware methods (library(Matrix)
# was attached above) and are much faster for larger matrices
table(rowSums(M4))
table(colSums(M4))

stopifnot(
  nrow(M4) == 188L,
  ncol(M4) == 282L,
  all(Matrix::rowSums(M4) == 6L),
  all(Matrix::colSums(M4) == 4L)
)

total_incidences <- sum(M4)
total_incidences
stopifnot(total_incidences == 188L * 6L, total_incidences == 282L * 4L)
```

## Pairwise overlap between specimens

The PP construction guarantees that any two distinct specimens share at most
$d - 1$ pools. With $d = 3$ here, that bound is 2. The specimen-by-specimen
overlap matrix is $\mathbf{M}^\top\mathbf{M}$; its off-diagonal entries are
the pairwise pool-sharing counts $|\mathcal{J}_i \cap \mathcal{J}_{i'}|$:

```{r overlap}
overlap <- Matrix::crossprod(M4)
diag(overlap) <- 0L  # ignore the diagonal (each specimen "overlaps" itself)

max_overlap <- max(overlap)
max_overlap

stopifnot(max_overlap <= 2L)  # the d - 1 = 2 theoretical bound
```

The maximum pairwise overlap actually attained, `r max_overlap`, is within
the theoretical bound $d - 1 = 2$.

# Efficiency

Testing all 282 specimens individually would need 282 tests. This design
needs only $J = 188$ pooled tests, a saving of

```{r efficiency}
n_specimens <- ncol(M4)
n_pools <- nrow(M4)

n_specimens
n_pools

saving_pct <- 100 * (1 - n_pools / n_specimens)
round(saving_pct, 1)
```

so pooling 282 specimens into 188 pools cuts the number of tests by about
`r round(saving_pct, 1)`%, at the cost of the resolution limits discussed
next.

# Decoding

`pp_decode()` runs the package's combined-oracle stage-1 pipeline (COMP, then
GPSR refinement if needed) on observed pool outcomes; see `?pp_decode` and
the `ppgt_pp` vignette for the full pipeline. This matrix's stored detection
limit is:

```{r kmax}
attr(M4, "pp")$k_max
```

so exact COMP recovery is guaranteed only up to `r attr(M4, "pp")$k_max`
true positive out of 282. The example below simulates exactly that many
positives with a fixed seed, so the whole example is reproducible.

```{r decode}
set.seed(20260926)

Y_tilde <- integer(n_specimens)
Y_tilde[137] <- 1L  # the one simulated positive specimen

# Latent pool status: z_tilde_j = 1 iff pool j contains at least one positive
z_tilde <- as.integer(as.vector(M4 %*% Y_tilde) > 0L)
sum(z_tilde)  # number of positive pools

decoded <- pp_decode(M4, z_tilde, verbose = FALSE)
decoded$positives
decoded$method

stopifnot(identical(decoded$positives, which(Y_tilde == 1L)))
```

`pp_decode()` recovers the single simulated positive specimen exactly, using
only the `r sum(z_tilde)` positive pools out of 188 as evidence.

# Matrix structure

```{r heatmap, fig.width=7, fig.height=5}
Matrix::image(
  M4,
  main = "188 x 282 PP pooling matrix (pools x specimens)",
  xlab = "Specimen",
  ylab = "Pool"
)
```

Each layer of `nl = 4` contributes a horizontal band of $q = 47$ pools; the
banded pattern reflects the polynomial-evaluation structure described in the
`ppgt_pp` vignette.

# Session info

```{r session-info}
sessionInfo()
```
